---
title: "Creating Tables - Tables"
header-includes:
   - \usepackage{xcolor}
   - \definecolor{umn}{HTML}{FF2D21}
   - \usepackage{caption}
   - \captionsetup[table]{textfont={it}, labelfont={bf}, singlelinecheck=false, labelsep=newline}
   - \captionsetup[figure]{textfont={}, labelfont={it}, singlelinecheck=false, labelsep=period}
   - \usepackage{xfrac}
output: 
  pdf_document:
    highlight: tango
    latex_engine: xelatex
    fig_width: 6
    fig_height: 6
mainfont: "Sabon"
sansfont: "Helvetica Neue UltraLight"
monofont: Inconsolata
urlcolor: "umn"
always_allow_html: yes
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Load libraries
library(corrr)
library(dplyr)
library(knitr)
library(kableExtra)
library(readr)
library(tidyr)

# Read in data
ed = read_csv(file = "~/Documents/github/epsy-8252/data/ed-schools-2018.csv") %>% drop_na()
riverview = read_csv(file = "~/Documents/github/epsy-8252/data/riverview.csv")
```


```{r echo=FALSE}
# Set up data frame
data.frame(
  Measure = c("Peer rating", "Acceptance rate for Ph.D. students", "Enrollment", "GRE score (verbal)", "GRE score (quantitative)"),
  M  = c(mean(ed$peer), mean(ed$doc_accept), mean(ed$enroll), mean(ed$gre_verbal), mean(ed$gre_quant)),
  SD = c(sd(ed$peer), sd(ed$doc_accept), sd(ed$enroll), sd(ed$gre_verbal), sd(ed$gre_quant))
) %>%
  kable(
    col.names = c("Measure", "\\textit{M}", "\\textit{SD}"),
    align = c("l", "c", "c"),
    digits = 1,
    caption = "\\\\[0.1ex]Means and Standard Deviations of Five Measures of Graduate Programs of Education ($N=122$)",
    format = "latex",
    booktabs = TRUE,
    escape = FALSE
    ) %>%
  kable_styling(latex_options = "HOLD_position", position = "left")
```


\vspace{1in}

```{r echo=FALSE}
# Set up data frame
data.frame(
  Measure = c("Education level (in years)", "Seniority (in years)", "Income (in U.S. dollars"),
  M_f  = c(
    mean(riverview$education[riverview$gender == "female"]),
    mean(riverview$seniority[riverview$gender == "female"]),
    mean(riverview$income[riverview$gender == "female"])
  ),
  SD_f = c(
    sd(riverview$education[riverview$gender == "female"]),
    sd(riverview$seniority[riverview$gender == "female"]),
    sd(riverview$income[riverview$gender == "female"])
    ),
  M_m  = c(
    mean(riverview$education[riverview$gender == "male"]), 
    mean(riverview$seniority[riverview$gender == "male"]),
    mean(riverview$income[riverview$gender == "male"])
    ),
  SD_m = c(
    sd(riverview$education[riverview$gender == "male"]), 
    sd(riverview$seniority[riverview$gender == "male"]),
    sd(riverview$income[riverview$gender == "male"])
    )
) %>%
  kable(
    col.names = c("Measure", "\\textit{M}", "\\textit{SD}", "\\textit{M}", "\\textit{SD}"),
    align = c("l", "c", "c", "c", "c"),
    digits = 0,
    caption = "\\\\[0.1ex]Means and Standard Deviations of Three Measures of Riverview Employees Conditioned on Sex",
    format = "latex",
    booktabs = TRUE,
    escape = FALSE
    ) %>%
  kable_styling(latex_options = "HOLD_position", position = "left") %>%
  add_header_above(
    header = c(" " = 1, "Females" = 2, "Males" = 2)
    )
```


\newpage

```{r echo=FALSE}
# Set up data frame
data.frame(
  Measure = c("Education level (in years)", "Seniority (in years)", "Income (in U.S. dollars"),
  M_f  = c("16 (4)", "14 (7)", "48938	(13265)"),
  CI_f = c("[13.7, 17.8]", "[10.7, 17.5]", "[42342, 55535]"),
  M_m  = c("16 (5)", "16 (7)", "59919	(14210)"),
  CI_m = c("[13.5, 19.1]", "[11.5, 19.9]", "[51714, 68123]")
) %>%
  kable(
    col.names = c("Measure", "\\textit{M} (\\textit{SD})", "95\\% CI", "\\textit{M} (\\textit{SD})", "95\\% CI"),
    align = c("l", "c", "c", "c", "c"),
    digits = 0,
    caption = "\\\\[0.1ex]Means, Standard Deviations, and Confidence Intervals (CIs) of Three Measures of Riverview Employees Conditioned on Sex",
    format = "latex",
    booktabs = TRUE,
    escape = FALSE
    ) %>%
  kable_styling(latex_options = "HOLD_position", position = "left")  %>%
  add_header_above(
    header = c(" " = 1, "Females" = 2, "Males" = 2)
  )
```


\vspace{1in}

```{r echo=FALSE, message=FALSE}
tab_04 = ed %>%
  select(peer, doc_accept, enroll, gre_verbal, gre_quant) %>%
  correlate() %>%
  shave(upper = TRUE) %>%
  fashion(decimals = 2, na_print = "—") %>%
  mutate(
    rowname = c("1. Peer rating", "2. Acceptance rate for Ph.D. students", "3. Enrollment", "4. GRE score (verbal)", "5. GRE score (quantitative)")
  )

tab_04[1, 3:6] = ""
tab_04[2, 4:6] = ""
tab_04[3, 5:6] = ""
tab_04[4, 6] = ""
    
tab_04 %>%
  kable(
    caption = "\\\\[0.1ex]Intercorrelations between Five Measures of Graduate Programs of Education",
    col.names = c("Measure", "1", "2", "3", "4", "5"),
    align = c("l", "c", "c", "c", "c", "c"),
    format = "latex",
    booktabs = TRUE,
    escape = FALSE
  ) %>%
  kable_styling(latex_options = "HOLD_position", position = "left")
```

\newpage

```{r echo=FALSE, message=FALSE}
tab_05_2 = ed %>%
  select(peer, doc_accept, enroll, gre_verbal, gre_quant) %>%
  correlate() %>%
  shave(upper = TRUE) %>%
  fashion(decimals = 2, na_print = "—") 

tab_05_1 = data.frame(
  M = c(3.3, 40.1, 970, 154.9, 151.0),
  SD = c(0.5, 20.2, 665, 3.7, 4.4)
)

tab_05 = cbind(tab_05_2[, 1], tab_05_1, tab_05_2[, 2:6])

tab_05[1, 5:8] = ""
tab_05[2, 6:8] = ""
tab_05[3, 7:8] = ""
tab_05[4, 8] = ""
  
tab_05 %>% 
  mutate(
    `tab_05_2[, 1]` = c("1. Peer rating", "2. Acceptance rate for Ph.D. students", "3. Enrollment", "4. GRE score (verbal)", "5. GRE score (quantitative)")
  ) %>%
  kable(
    caption = "\\\\[0.1ex]Intercorrelations between Five Measures of Graduate Programs of Education. Means and Standard Deviations are Presented Along the Main Diagonal",
    col.names = c("Measure", "\\textit{M}", "\\textit{SD}", "1", "2", "3", "4", "5"),
    align = c("l", "c", "c", "c", "c", "c", "c", "c"),
    format = "latex",
    booktabs = TRUE,
    escape = FALSE
  ) %>%
  kable_styling(latex_options = "HOLD_position", position = "left")
```


\vspace{1in}


```{r echo=FALSE}
tab_06 = lm(peer ~ 1 + doc_accept + enroll + gre_verbal + gre_quant, data = ed) %>%
  broom::tidy() %>%
  mutate(
    term = c("Constant", "Acceptance rate for Ph.D. students", "Enrollment", "GRE score (verbal)", "GRE score (quantitative)")
  )

tab_06 = tab_06[c(2:5, 1), ]

tab_06 %>%
  kable(
    caption = "\\\\[0.1ex]Coefficient-Level Estimates for a Model Fitted to Estimate Variation in Peer Ratings",
    col.names = c("Predictor", "\\textit{B}", "\\textit{SE}", "\\textit{t}", "\\textit{p}"),
    digits = c(0, 2, 3, 2, 3),
    align = c("l", "c", "c", "c", "c"),
    format = "latex",
    booktabs = TRUE,
    escape = FALSE
  ) %>%
  kable_styling(latex_options = "HOLD_position", position = "left")
```


\newpage


```{r echo=FALSE}
tab_06 %>%
  mutate(
    CI = c("[-0.014, -0.006]", "[0.000, 0.000]", "[-0.031, 0.034]", "[0.007, 0.061]", "[-4.894, 1.521]"),
    Beta = c(-.43, .07, .01, .32, .01)
  ) %>%
  select(term, estimate, std.error, CI)  %>%
  kable(
    caption = "\\\\[0.1ex]Coefficient-Level Estimates and Uncertainty for a Model Fitted to Estimate Variation in Peer Ratings",
    col.names = c("Predictor", "\\textit{B}", "\\textit{SE}", "95\\% CI"),
    digits = c(0, 2, 2, 0),
    align = c("l", "r", "c", "c"),
    format = "latex",
    booktabs = TRUE,
    escape = FALSE
  ) %>%
  kable_styling(latex_options = "HOLD_position", position = "left")
```


\vspace{1in}


```{r echo=FALSE}
tab_06 %>%
  mutate(
    CI = c("[-0.014, -0.006]", "[0.000, 0.000]", "[-0.031, 0.034]", "[0.007, 0.061]", "[-4.894, 1.521]"),
    Beta = c(-.43, .07, .01, .32, .01)
  ) %>%
  select(term, estimate, std.error, CI, Beta)  %>%
  kable(
    caption = "\\\\[0.1ex]Coefficient-Level Estimates (Unstandardized and Standardized) and Uncertainty for a Model Fitted to Estimate Variation in Peer Ratings",
    col.names = c("Predictor", "\\textit{B}", "\\textit{SE}", "95\\% CI", "$\\beta$"),
    digits = c(0, 2, 2, 0, 2),
    align = c("l", "c", "c", "c", "c"),
    format = "latex",
    booktabs = TRUE,
    escape = FALSE
  ) %>%
  kable_styling(latex_options = "HOLD_position", position = "left")
```


\newpage


```{r, message=FALSE, echo=FALSE, results='asis', eval=FALSE}
library(stargazer)

lm.1 = lm(peer ~ 1 + gre_verbal + gre_quant, data = ed)
lm.2 = lm(peer ~ 1 + doc_accept + enroll, data = ed)
lm.3 = lm(peer ~ 1 + doc_accept + enroll + gre_verbal + gre_quant, data = ed)

stargazer(lm.1, lm.2, lm.3,
  ci = TRUE,
  column.labels = c("Model 1", "Model 2", "Model 3"),
  covariate.labels = c("GRE score (verbal)", "GRE score (quantitative)", "Acceptance rate for Ph.D. students", "Enrollment"),
  dep.var.caption = "Outcome variable: Peer ratings",
  dep.var.labels = NULL,
  dep.var.labels.include = FALSE,
  type = "latex",
  keep.stat = c("rsq","ser"),
  star.cutoffs = NA,
  header = FALSE,
  table.placement = 'H',
  title = "Unstandardized Coefficients and Confidence Intervals for a Series of Regression Models Fitted to Data from $n=129$ Graduate Schools of Education to Predict Variation in Peer Ratings",
  omit.table.layout = "n"
  )
```

\begin{table}[H]  
  \caption{Unstandardized Coefficients and Confidence Intervals for a Series of Regression Models Fitted to Data from $n=129$ Graduate Schools of Education to Predict Variation in Peer Ratings} 
  \label{} 
\begin{tabular}{@{\extracolsep{5pt}}lccc} 
\\[-1.8ex]\hline\\[1ex] 
 & Model 1 & Model 2 & Model 3 \\ 
\hline \\[-1.8ex] 
 GRE score (verbal) & 0.011 &  & 0.001 \\ 
  & ($-$0.024, 0.046) &  & ($-$0.031, 0.033) \\ 
  & & & \\ 
 GRE score (quantitative) & 0.047 &  & 0.036 \\ 
  & (0.017, 0.076) &  & (0.009, 0.063) \\ 
  & & & \\ 
 Acceptance rate for Ph.D. students &  & $-$0.013 & $-$0.010 \\ 
  &  & ($-$0.017, $-$0.009) & ($-$0.014, $-$0.006) \\ 
  & & & \\ 
 Enrollment &  & 0.0001 & 0.0001 \\ 
  &  & ($-$0.00004, 0.0002) & ($-$0.0001, 0.0002) \\ 
  & & & \\ 
 Constant & $-$5.488 & 3.769 & $-$1.857 \\ 
  & ($-$8.683, $-$2.294) & (3.572, 3.967) & ($-$5.054, 1.340) \\ 
  & & & \\ 
\hline \\[-1.8ex] 
R$^{2}$ & 0.243 & 0.300 & 0.390 \\ 
RMSE & 0.429 & 0.413 & 0.389 \\ 
\hline 
\end{tabular} 
\end{table} 



\newpage

```{r, message=FALSE, echo=FALSE, results='asis', eval=FALSE}
stargazer(lm.1, lm.2, lm.3,
  column.labels = c("Model 1", "Model 2", "Model 3"),
  covariate.labels = c("GRE score (verbal)", "GRE score (quantitative)", "Acceptance rate for Ph.D. students", "Enrollment"),
  dep.var.caption = "Outcome variable: Peer ratings",
  dep.var.labels = NULL,
  dep.var.labels.include = FALSE,
  type = "latex",
  keep.stat = c("rsq","ser"),
  report = c("vcsp"),
  star.cutoffs = NA,
  header = FALSE,
  table.placement = 'H',
  title = "Unstandardized Coefficients and Confidence Intervals for a Series of Regression Models Fitted to Data from $n=129$ Graduate Schools of Education to Predict Variation in Peer Ratings",
  omit.table.layout = "n"
  )
```



\begin{table}[H] 
  \caption{Unstandardized Coefficients and Confidence Intervals for a Series of Regression Models Fitted to Data from $n=129$ Graduate Schools of Education to Predict Variation in Peer Ratings} 
  \label{} 
\begin{tabular}{@{\extracolsep{5pt}}lccc} 
\\[-1.8ex]\hline\\[1ex] 
 & Model 1 & Model 2 & Model 3 \\ 
\hline \\[-1.8ex] 
 GRE score (verbal) & 0.011 &  & 0.001 \\ 
  & (0.018) &  & (0.016) \\ 
  & $p = 0.531$ &  & $p = 0.950$ \\ 
  & & & \\ 
 GRE score (quantitative) & 0.047 &  & 0.036 \\ 
  & (0.015) &  & (0.014) \\ 
  & $p = 0.003$ &  & $p = 0.012$ \\ 
  & & & \\ 
 Acceptance rate for Ph.D. students &  & $-$0.013 & $-$0.010 \\ 
  &  & (0.002) & (0.002) \\ 
  &  & $p = 0.000$ & $p = 0.00000$ \\ 
  & & & \\ 
 Enrollment &  & 0.0001 & 0.0001 \\ 
  &  & (0.0001) & (0.0001) \\ 
  &  & $p = 0.239$ & $p = 0.347$ \\ 
  & & & \\ 
 Constant & $-$5.488 & 3.769 & $-$1.857 \\ 
  & (1.630) & (0.101) & (1.631) \\ 
  & $p = 0.002$ & $p = 0.000$ & $p = 0.258$ \\ 
  & & & \\ 
\hline \\[-1.8ex] 
R$^{2}$ & 0.243 & 0.300 & 0.390 \\ 
RMSE & 0.429 & 0.413 & 0.389 \\ 
\hline 
\hline \\[-1.8ex] 
\end{tabular} 
\end{table} 


\newpage

\begin{table}[H] 
  \caption{Unstandardized Coefficients and Confidence Intervals for a Series of Regression Models Fitted to Data from $n=129$ Graduate Schools of Education to Predict Variation in Peer Ratings} 
  \label{} 
\begin{tabular}{@{\extracolsep{5pt}}lccc} 
\\[-1.8ex]\hline\\[1ex] 
 & Model 1 & Model 2 & Model 3 \\ 
\hline \\[-1.8ex] 
 GRE score (verbal) & 0.011 (0.018) &  & 0.001 (0.016) \\ 
  & $p = 0.531$ &  & $p = 0.950$ \\ 
  & & & \\ 
 GRE score (quantitative) & 0.047 (0.015) &  & 0.036 (0.014) \\ 
  & $p = 0.003$ &  & $p = 0.012$ \\ 
  & & & \\ 
 Acceptance rate for Ph.D. students &  & $-$0.013 (0.002) & $-$0.010 (0.002) \\ 
  &  & $p = 0.000$ & $p = 0.00000$ \\ 
  & & & \\ 
 Enrollment &  & 0.0001 (0.0001) & 0.0001 (0.0001)\\ 
  &  & $p = 0.239$ & $p = 0.347$ \\ 
  & & & \\ 
 Constant & $-$5.488 (1.630) & 3.769 (0.101) & $-$1.857 (1.631) \\ 
  & $p = 0.002$ & $p = 0.000$ & $p = 0.258$ \\ 
  & & & \\ 
\hline \\[-1.8ex] 
R$^{2}$ & 0.243 & 0.300 & 0.390 \\ 
RMSE & 0.429 & 0.413 & 0.389 \\ 
\hline 
\hline \\[-1.8ex] 
\end{tabular} 
\end{table} 


