## comic-characters.csv {-}

```{r}
#| echo: false
source("../scripts/_common.R")
```

The data in **comic-characters.csv** contains data on 14 attributes for 23,272 comic characters. These data were scraped in 2014 from from [Marvel Wikia](http://marvel.wikia.com/Main_Page) and [DC Wikia](http://dc.wikia.com/wiki/Main_Page) by FiveThirtyEight and used in the story [Comic Books Are Still Made By Men, For Men And About Men](http://fivethirtyeight.com/features/women-in-comic-books/). The variables are:

- `character`: The name of the character
- `comic`: If the character appears in DC Comics or Marvel Comics
- `reality`: Comic reality the character appears in
- `identity`: The identity status of the character (Secret Identity, Public identity, No Dual Identity)
- `alignment`: If the character is Good, Bad or Neutral
- `eye_color`: Eye color of the character
- `hair_color`: Hair color of the character 
- `sex`: Sex of the character (e.g. Male, Female, etc.)
- `lgbtq`: If the character is identified as LGBTQ
- `lgbtq_note`: Additional information if the character is identified as LGBTQ
- `alive`: If the character is alive or deceased
- `appearances`: The number of appearances of the character in comic books (as of September 2, 2014)
- `first_appear_date`: The month and year of the character's first appearance in a comic book, if available
- `first_appear_year`: The year of the character's first appearance in a comic book, if available

<br />


### Preview

```{r}
#| code-fold: true
# Import Data
comics = readr::read_csv(file = "https://raw.githubusercontent.com/zief0002/modeling/main/data/comic-characters.csv")

# View data
comics
```

<br />



