## state-education.csv {-}

```{r}
#| echo: false
source("../scripts/_common.R")
```

This data, stored in **state-education.csv** includes state-level aggregate data on six attributes from academic year 2020--21. The attributes, collected for all 50 states and the District of Columbia, are:

- `state`: State name
- `postal`: State postal code
- `region`: Region of the country identified by the National Education Association (Far West, Great Lakes, Mid East, New England, Plains, Rocky Mountains, Southeast, Southwest)
- `salary`: Average public teacher salary in the state.
- `sat`: Average total SAT score in the state for students in the class of 2020 who took the SAT during high school. The SAT is made up of three sections: (1) Reading, (2) Writing and Language (also just called Writing), and (3) Math. Total SAT score range from 400--1600.
- `participation`: Percentage of students in the state who took the SAT.



<br />


### Preview

```{r}
#| code-fold: true
# Import Data
state_educ = readr::read_csv(file = "https://raw.githubusercontent.com/zief0002/modeling/main/data/state-education.csv")

# View Data
state_educ
```



<br />


### References

College Board. (2020). [SAT suite of assessments annual report](https://reports.collegeboard.org/media/pdf/2020-total-group-sat-suite-assessments-annual-report.pdf). Author.

NEA Research. (2021). [Rankings of the states 2020 and estimates of school statistics 2021](https://www.nea.org/sites/default/files/2021-04/2021%20Rankings_and_Estimates_Report.pdf). National Education Association.


<br />


