## gapminder.csv {-}

```{r}
#| echo: false
source(",,/scripts/_common.R")
```

The data in **gapminder.csv** contains data on 6 attributes for 193 countries. These data, based on 2017 measures collected by World Bank, were compiled by [gapminder.org](https://www.gapminder.org/) and made available free under the [CC-BY license](https://creativecommons.org/licenses/by/4.0/). The variables are:

- `country`: The name of the country
- `region`: The world region (Africa, Americas, Asia, or Europe)
- `income`: Income per person based on the gross domestic product per person, adjusted for differences in purchasing power (in international dollars, fixed for 2017 prices)
- `life_exp`: Average number of years a newborn child would live if current mortality patterns were to stay the same
- `co2`: Carbon dioxide emissions per person from the burning of fossil fuels (metric tonnes of CO2 per person)
- `co2_change`: Indicator of whether the carbon dioxide emissions per person from the burning of fossil fuels has increased or decreased since 2007 (10 year span)
- `population`: Population of the country, in millions



<br />


### Preview

```{r}
#| code-fold: true
# Import Data
gapminder = readr::read_csv(file = "https://raw.githubusercontent.com/zief0002/modeling/main/data/gapminder.csv")

# View data
gapminder
```

<br />


### References

College Board. (2020). [SAT suite of assessments annual report](https://reports.collegeboard.org/pdf/2020-total-group-sat-suite-assessments-annual-report.pdf). Author.

NEA Research. (2021). [Rankings of the states 2020 and estimates of school statistics 2021](https://www.nea.org/sites/default/files/2021-04/2021%20Rankings_and_Estimates_Report.pdf). National Education Association.

