## work-demands.csv {#work}

```{r}
#| echo: false
source("scripts/_common.R")
```

Research has documented the importance of a healthy work-life balance. The data in [work-demands.csv](https://raw.githubusercontent.com/zief0002/modeling/main/data/work-demands.csv) were simulated to mimic the effects of receiving work-related contact outside of normal work hours found in the literature on psychological outcomes. The variables in the data are:

- `guilt`: Measure of guilt (as a transitory affective state) related to home-life balance. This was measured using survey questions such as,  "In the past seven days, on how many  days  have  you  felt  guilty?" Higher values indicate more perceived guilt.
- `contact`: Measure of work contact/demands outside of normal working hours. Measured using survey items such as, "how  often  do  coworkers,  supervisors, managers, customers, or clients contact you about work-related  matters  outside  normal  work  hours?". Higher values indicate higher degrees of boundary-spanning work demands.
- `female`: Dummy-coded indicator of sex (0 = Not female; 1 = Female)
- `authority`: Standardized measure of job authority. Measured using survey items such as: "Do you influence or set the rate of pay  received  by  others?" and "Do you have  the authority to hire or fire others?". Higher values indicate more authority.
- `married`: Dummy-coded indicator of marital status (0 = Not married; 1 = Married)



<br />


### Preview

```{r}
#| code-fold: true
# Import Data
work = readr::read_csv(file = "https://raw.githubusercontent.com/zief0002/modeling/main/data/work-demands.csv")

# View data
work
```

<br />


### References


Glavin, P., Schieman, S., &amp; Reid, S. (2011). Boundary-spanning work demands and their consequences for guilt and psychological distress. *Journal of Health and Social Behavior, 52*(1) 43&ndash;57. doi: 10.1177/0022146510395023

